An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-18 manuscript.Rmd
updated references, formatting to acm style, minor textual edits

---
layout: preprint
title: "An introduction to Docker for reproducible research, with examples from the R environment"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
abstract: |
  Docker is a promising solution to

bibliography: components/references.bib
csl: components/ecology.csl
documentclass: components/elsarticle

output: 
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---

@@

  * novel ways to manage systems research artifacts, dependencies, workflows,
    and data collection
  * testbeds for repeatable research and/or publishing executable artifacts
  * "rich publications" that connect papers to software and data
  * repeatability in distributed and nondeterministic systems
  * mitigating obstacles to artifact sharing, repeatability, and reproduction
  * repeatability challenges specific to systems
  * measurement bias in systems experiments
  * experiences and outcomes from systems repetition and reproduction studies
  * improving scientific practice in systems research
  * incentives for producing and sharing high-quality research artifacts

Papers are expected to present significant results, insights, and/or directions
for future work, and they are expected to include substantial material that has
not already been published.

SUBMISSION INSTRUCTIONS

Submissions must be no longer than ten (10) pages; shorter papers are allowed
and encouraged.  Submissions must be written in English and follow the standard
formatting guidelines for papers appearing in Operating Systems Review (see
<http://www.sigops.org/osr.html>).  Papers must be submitted in PDF format via
the submission web site.  Submitted papers will be reviewed by the guest editor
and the review committee.  Papers will be evaluated on technical quality,
originality, relevance, and presentation.  Accepted papers will be published in
the January 2015 issue of Operating Systems Review.

Reproducible research has recieved an increasing level of attention
throughout the scientific community (e.g. [Peng 2011], [Nature 2014])
and the public at large (e.g. [The Economist 2013]). All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has recieved particular
attention [Merali 2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

# A cultural problem

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place ([Barnes 2010]). Despite extensive evidence to the contrary
(e.g. [Ince 2012]), many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, we cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

A lack of requirements or incentives no doubt plays a crucial role in
discouraging sharing ([Barnes 2010], [Stodden 2013]). Nevertheless,
it is easy to underestimate the siginficant barriers raised by a lack
of familiar, intuitive, and widely adopted tools for addressing the
challenges of computational reproducibility. Surveys and case studies find
that a lack of time, more than innate opposition to sharing, discourages
researchers from providing code ([Stodden 2010], [FitzJohn 2014]).

# The four technical challenges of computational reproducibility

By restricting ourselves to studies of where code has been made
available, we may sidestep for the moment the cultural challenges
to reproducibility so that we may focus on the technical ones; in
particular, those challenges for which improved tools and techniques
rather than merely norms of behavior can contribute substantially to
improved reproducibility.

Studies focusing on code that has been made available with scientific
publications regularly find the same common issues that pose substantial
barriers to reproducing the original results or building on that code
(e.g. [Lapp 2014], [Garijo 2013], [Gilbert 2012], [Collberg 2014],
[Brown 2014]), which I attempt to summarize as follows.

<!-- Is this a permitted term? -->

1. **"Dependency Hell"**

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[Collberg 2014], and in an ongoing effort by other researchers to
replicate that study [Brown 2014]. Installing or building software
necessary to run the code in question assumes the ability to recreate
the computational environment of the original researchers.

<!-- Time savings. Personal reproducibility. Collaboration. Teaching. -->

<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents
-->

<!-- Omit this? -->
Differences in numerical evaluation, such as arise in floating point
arithmetic or computers or even ambiguities in standardized programming
languages ("order-of-evaluation" problems) can be responsible for
differing results between or even within the same computational platform
([Ince 2012]). Such issues make it difficult to restrict the true dependencies
of the code to higher level environments such as that of a given scripting
language, indpendent of the underlying OS or even hardware itself.

2.  **Imprecise documentation**

Documentation on how to install and run code associated with published
research is another frequent barrier to replication.  [Lapp 2014] found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" (Sensu [Garijo 2013] --
where novices may be experts in nearby langauages but unfamiliar with
the package managers and other tools of the language involved). This
same problem is discussed in [Clark 2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
(n=34) using the popular software STRUCTURE could be reporoduced in the
study of [Gilbert 2012].

3.  **Code rot**

Software dependencies are not static elements, but recieve regular
updates that may fix bugs, add new features or deprecate old features
(or even entire dependencies themselves).  Any of these changes can
potentially change the results generated by the code.  As some of these
changes may indeed resolve valid bugs or earlier problems with underlying
code, it will often be insufficient to demonstrate that results can be
reproduced when using the original versions, a problem sometimes known
as "code rot."  Researchers will want to know if the results are robust
to the changes. The case studies in [Lapp 2014] provide examples of
these problems.

4. **Barriers to adoption and reuse in existing solutions**

Technological solutions such as workflow software, virtual machines,
continuous integration services, and best practices from software
development would address many of the issues frequently frustrating
reproducibility. However, researchers face siginificant barriers to entry
in learning these tools and approaches which are not part of their typical
curriculum, or lack incentives co-measurate with the effort required
([Joppa 2013], [FitzJohn 2014]).

@@

<!-- A wide range of proposals, we focus on those at the operating system level -->
Though a wide variety of approaches exists to work around these challenges,
few operate on a low enough level to provide a general solution.  [Clark 2014]
provide an excellent description of this situation:

> In scientific computing the environment was commonly managed via
Makfiles & Unix-y hacks, or alternatively with monolithic software
like Matlab. More recently, centralized package management has provided
curated tools that work well together. But as more and more essential
functionality is built out across a variety of systems and languages,
the value -- and also the difficulty -- of coordinating multiple tools
continues to increase. Whether we are producing research results or web
services, it is becoming increasingly essential to set up new languages,
libraries, databases, and more.

@@

Two dominant paradigms have emerged to address these issues so far:
workflow software (e.g. [Atintas 2004], [Hull 2006])  and
cloud computing on virtual machines (e.g. [Dudley 2013], [Howe 2012]). 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from of
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they recieve substatianal adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall ([Gin 2007, Dudley 2013]).

[Dudley 2013] gives several reasons that such comprehensive workflow
systems have not been more successful:

@@

In short, workflow software expects a new approach to computational research. 
In contrast, virtual machines (VMs) offer a more direct approach. Since
the computer Operating System (OS) already provides the software layer
responsible for coodinating all the different elements running on the computer,
the VM approach captures the OS and everything running on it whole-cloth.
To make this practical, [Dudley 2013] and [Howe 2012]
propose using virtual machine images that will run on the cloud, such as Amazon's
EC2 system, which is already based upon this kind of virtualization.

Critics of the use of VMs to support reproducibility highlight that the 
approach is too much of a black box and thus ill suited for reproducibility [Watson 2012].
While the approach sidesteps the need to either install or even document the
dependencies, this also makes it more difficult for other researchers to
understand, evaluate, or alter those dependencies.  Moreover, other research
cannot easily build on the virtual machine in a consistent and scalable way. 
If each study provided it's own virtual machine, any pipeline combining the 
tools of multiple studies would quickly become impractical or impossible 
to implement.

@@

The problems highlighted here are not unique to _academic_ software,
but impact software development in general. While the academic research
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual
machines, the software development community has recently emphasized a
philosophy (rather than a particular tool), known as _Development_ and
Systems _Operation_, or more frequently just "DevOpts." The approach is
characterized by scripting, rather than documenting, a description of the
necessary dependencies for software to run, usually from the Operating
System (OS) on up.  [Clark 2014] describes the DevOpts approach along
with both its relevance to reproducible research and examples of its
use in the academic research context. They identify the difficulties
we have discussed so far in terms of effective documentation:

@@

The authors contrast this to the DevOpts approach, where dependency 
documentation is _scripted_:

@@

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. [Clark 2014] are careful to note
that this is not so much a technological shift as a philosophical one:

@@

Nevertheless, a growing suite of tools designed explicitly for this
purpose have rapidly replaced the use of general purpose tools (such
as Makefiles, bash scripts) to become synonomous with the "DevOpts"
philosophy.  [Clark 2014] reviews many of these "DevOpts" tools, their
different roles, and their application in reproducible research.

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play
in reproducible research. Docker offers several promising features
for reproducibility that go beyond the tools highlighted in [Clark
2014]. Nevertheless, my goal in focusing on this technology is not to
promote a particular solution, but to anchor the discussion of technical
solutions to reproducibility challenges in concrete examples.

@@

Docker is an open source project that builds on many long-familiar
technologies from operating systems research: LXC containers,
virtualization of the OS, and a hash-based or git-like verioning
and differencing system, among others.

<!-- Define:
- Dockerfile
- Image 
- Container: When we run a docker image, we get a container.  We can run the same image multiple times, resulting in many identical containers.  We can save the state of a container at any time, which creates a new image.  
- Dockerfile -> build -> image
- Image -> run -> container
-->

@@

A key difference between Docker images and other virutal machines is 
that the Docker images share the Linux kernel with the host machine.
For the end user the primary consequence of this is that any Docker
image must be based on a Linux system with Linux-compatible software, which
includes (R, Python, Matlab, and most other scientific programming needs). [^4]

Sharing the Linux kernel makes Docker much more light-weight and higher
performance than complete virtual machines -- a typical desktop computer
could run no more than a few virutal machines at once but 
would have no trouble running 100s of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

@@

Though Docker images can be created interactively, this leaves little
transparent record [^3] of what software has been installed and how.
Dockerfiles provide a simple script (similar to a Makefile) that defines
exactly how to build up the image, consistent with the DevOpts approach
we mentioned previously.

With a syntax that is simpler than other provisioning tools (e.g. Chef,
Puppet, Ansible) or Continuous Integration (CI) platforms (e.g. Travis CI,
Shippable CI); users need little more than a basic familarity with shell
scripts and a Linux distribution software environment (e.g. Debian-based
`apt-get`) to get started writing Dockerfiles.

@@

- the `Dockerfile` provides a human readable summary of the necessary
software dependencies, environmental variables and so forth needed to
execute the code. There is little possiblity of the kind of holes or 
imprecision in such a script that so frequently cause difficulty in
manually implemented documentation of dependencies. This approach
also avoids the burden of having to tediously document dependencies 
at the end of a project, since they are instead documented as they 
are installed by writing the `Dockerfile`.

- Unlike a `Makefile` or other script, the `Dockerfile` includes all
software dependencies down to the level of the OS, and is built by the
Docker `build` tool, making it very unlikely that the resulting build
will differ when being built on different machines. This is not to say
that all builds of a Dockerfile are bitwise identical.  In particular,
builds executed later will install more recent versions of the same
software, if available, unless the package managers used are explicitly
configured otherwise. We address this issue in the next section.

- It is possible to add checks and tests following the commands for
installing the software environment, which will verify that the setup
has been successful.  This can be important in addressing the issue of
code-rot which we discuss next.

@@

[^3]: The situation is in fact slightly better than the virtual machine
approach because these changes are versioned. Docker provides tools
to inspect differences (diffs) between the images, and we can also roll
back changes to earlier versions.

@@

As I have discussed above, changes to the dependencies, whether they are
the result of security fixes, new features, or deprecation of old software,
can break otherwise functioning code. These challenges can be significantly
reduced because Docker defines the software environment to a particular operating system 
and suite of libraries, such as the Ubuntu or Debian distribution.  Such 
distributions use a staged release model with `stable`, `testing` and `unstable` 
phases subjected to extensive testing to catch such potential problems [Ooms 2013],
while also providing regular security updates to software within each stage.
Nonetheless, this cannot completely avoid the challenge of code-rot,
particularly when it is necessary to install software that is not (yet)
available for a given distribution.

@@

<!-- FIXME move this later --> 
To simplify this process, Docker also supports *Automated Builds*
through the Docker Hub.  This acts as a kind of Continous Integration
(CI) service that verifies the image builds correctly whenever the
Dockerfile is updated, particularly if the Dockerfile includes checks
for the environment. The Hub also provides a convenient distribution
service, freely storing the prebuilt images, along with their metadata,
for download and reuse by others. The Docker Hub is a free service and
an open source software product so that users can run their own private
versions of the Hub on their own servers, for instance, if security of
the data or the longevity of the public platform is a concern.

@@

Though most of the concerns we have discussed so far can be addressed
through well-designed workflow software or the use of a DevOpts approach
to provisioning virtual machines by scripts, neither approach has seen
widespread adoption by domain researchers, who work primarily in a
local rather than cloud-based environment using develoment tools native
to their personal operating system. To gain more widespread adoption,
reproducible research technologies must make it easier, not harder,
for a researcher to perform the tasks they are already doing (before 
considering any additional added benefits).

These issues are reflected both during the original research or development
phase and in any subsequent reuse. Another researcher may be less likely
to build on existing work if it can only be done by using a particular
workflow system or monolothic software platform with which they are
unfamiliar.  Likewise, a user is more likey to make their own computational
environment avialable for reuse if it does not involve a significant
added effort in packaging and documenting ([Stodden 2010]).

Though Docker is not immune to these challenges, it offers a interesting 
example of a way forward in addressing these fundamental cocerns.

@@

# Using Docker as a local development environment

Perhaps the most important feature of a reproducible research tool is
that it be easy to learn and fit relatively seamlessly into exsiting
workflow patterns of domain researchers.  As we have seen, this,
more than any other concern, can explain the relatively low uptake
of previously proposed solutions.  Being a new and unfamiliar tool to
most domain scientists, Docker is far from immune to the same critique.
Nevertheless, Docker takes us several key steps towards an approach that
can be easily adopted in a research context.

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments (e.g. [Dudley 2014]), many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces are rapidly
filling this gap).

@@

For example, we launch an interactive R console in a container that
is linked to our current working directory:

```bash
docker run -v $(pwd):/ -it eddelbuettel/docker-ubuntu-r /usr/bin/R
```

@@

That will take a while to download the image the first time you run
it. boot2docker users will then need to determine the ip address assigned
to boot2docker, while Linux users can just specify `http://localhost`

@@

The key point here is that other researchers can easily build off this 
new image we have just created, extending our work directly, rather than
having to go back to the original image.

However, before others can re-use our image we need a way to share it.

### Docker hub

>>>>

---
title: "An introduction to Docker for reproducible research, with examples from the R environment"
author: 
  - name: Carl Boettiger
    affiliation: 
      l1: Center for Stock Assessment Research, 
      l2: 110 Shaffer Rd, Santa Cruz, CA 95050, USA 
    email: cboettig(at)gmail.com
    footnote: Corresponding author
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
abstract: |
  As computational work becomes more and more integral to many aspects
  of scientific research, computational reproducibility has become an
  issue of increasing importance to computer systems researchers and
  domain scientists alike. Though computational reproducibility seems 
  more straight forward than replicating physical experiments, the 
  immense, rapidly growing and frequently changing nature of computer
  environments makes being able to reproduce work in a way that others
  can understand and extend it a serious challenge.  In this paper, I
  explore common reasons that code developed for a gien research project
  cannot be successfully executed or extended by subsequent researchers.
  I review current approaches to these issues, including virtual machines
  and workflow systems, and their limitations. I then examine how the
  popular emerging technology Docker combines several areas from systems
  research - suc has operating system virtualization, cross-platform
  portability, modular re-usable elements, versioning, and a 'DevOps'
  philosophy, to address these challenges.  I illustrate this with several
  examples of Docker use with a focus on the R statistical environment.

Keywords:
  - docker
  - virtualization
  - reproducibility

bibliography: components/references.bib
csl: components/acm-sig-proceedings.csl

output: 
  pdf_document:
    template: components/sigproc-sp.tex
    keep_tex: true
    fig_caption: true
---

<!--
There is increasing interest in promoting repeatable and reproducible research
in computer science.  A community gains confidence in scientific claims when
the experiments supporting those claims are published, examined, and repeated.
Repeating research also helps a community expand on previous results and
discover new directions for investigation.  Repetition requires that the
artifacts used in an experiment be made available to, and made usable by,
investigators.  In other words, artifacts must be shared.  Possible artifacts
include not only binaries, source code, and datasets but also computing and
networking infrastructure such as testbeds and clouds.

Systems researchers face special challenges in the areas of repeatability and
artifact sharing.  Some arise from the nature of systems software, which is
typically complex and rife with internal and external dependencies.  Other
issues arise from execution environments: for example, to reproduce performance
and scalability results, one may need access to actual and specific hardware.
Still more systems-specific challenges arise from the difficulty of measuring
properties of interest, such as time and energy consumption.

@@

  * novel ways to manage systems research artifacts, dependencies, workflows,
    and data collection
  * testbeds for repeatable research and/or publishing executable artifacts
  * "rich publications" that connect papers to software and data
  * repeatability in distributed and nondeterministic systems
  * mitigating obstacles to artifact sharing, repeatability, and reproduction
  * repeatability challenges specific to systems
  * measurement bias in systems experiments
  * experiences and outcomes from systems repetition and reproduction studies
  * improving scientific practice in systems research
  * incentives for producing and sharing high-quality research artifacts
-->

# Introduction

Reproducible research has received an increasing level of attention
throughout the scientific community [e.g. @Peng_2011; @Nature_2012]
and the public at large [e.g. @The_Economist_2013]. All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has received particular
attention [@Merali_2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

### Systems research & reproducibility

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives [e.g. @Eide_2010, @Harji_2013]. Docker is a new but already
very popular open source tool that combines many of these approaches
in a user friendly implementation, including: (1) performing Linux
container (LXC) based operating system (OS) level virtualization, (2)
portable deployment of containers across platforms, (3) component reuse,
(4) sharing, (5) archiving, and (6) versioning of container images. While
Docker's market success has largely focused on the needs of businesses
in deploying web applications and the potential for a lightweight
alternative to full virtualization, these features have potentially
important implications for systems research in the area of scientific
reproducibility. In this paper, I seek to set these issues in the context
of reproducibility throughout the various domain sciences where 
computation plays an ever-increasing role, but where researchers are
largely unaware of the concerns or technologies involved in making these
computations more reproducible, extensible, and portable to other researchers. 
In so doing, I highlight elements of the Docker platform that should be
of interest to both computer systems research and domain scientists.

<!-- Map here! -->

### A cultural problem

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place [@Barnes_2010]. Despite extensive evidence to the contrary
[e.g. @Ince_2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, one cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

A lack of requirements or incentives no doubt plays a crucial role in
discouraging sharing [@Barnes_2010; @Stodden_2013]. Nevertheless,
it is easy to underestimate the significant barriers raised by a lack
of familiar, intuitive, and widely adopted tools for addressing the
challenges of computational reproducibility. Surveys and case studies find
that a lack of time, more than innate opposition to sharing, discourages
researchers from providing code [@Stodden_2010; @FitzJohn_2014].

## Four technical challenges

By restricting ourselves to studies of where code has been made
available, I will sidestep for the moment the cultural challenges
to reproducibility so that I may focus on the technical ones; in
particular, those challenges for which improved tools and techniques
rather than merely norms of behavior can contribute substantially to
improved reproducibility.

Studies focusing on code that has been made available with scientific
publications regularly find the same common issues that pose substantial
barriers to reproducing the original results or building on that code
[e.g. @Lapp_2014; @Garijo_2013; @Gilbert_2012; @Collberg_2014;
@Krishnamurthi_2014], which I attempt to summarize as follows.

### 1. **"Dependency Hell"**

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[@Collberg_2014], and in an ongoing effort by other researchers to
replicate that study [@Krishnamurthi_2014]. Installing or building software
necessary to run the code in question assumes the ability to recreate
the computational environment of the original researchers. 
<!-- Time savings. Personal reproducibility. Collaboration. Teaching. --> 
<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents
-->
<!-- Omit this? -->

Differences in numerical evaluation, such as arise in floating point
arithmetic or computers or even ambiguities in standardized programming
languages ("order-of-evaluation" problems) can be responsible for
differing results between or even within the same computational platform
[@Ince_2012]. Such issues make it difficult to restrict the true dependencies
of the code to higher level environments such as that of a given scripting
language, independent of the underlying OS or even hardware itself.

### 2.  **Imprecise documentation**

Documentation on how to install and run code associated with published
research is another frequent barrier to replication.  @Lapp_2014 found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" [Sensu @Garijo_2013] --
where novices may be experts in nearby languages but unfamiliar with
the package managers and other tools of the language involved). This
same problem is discussed in [@Clark_2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
($n=34$) using the popular software STRUCTURE could be reproduced in the
study of [@Gilbert_2012].

### 3.  **Code rot**

Software dependencies are not static elements, but receive regular
updates that may fix bugs, add new features or deprecate old features
(or even entire dependencies themselves).  Any of these changes can
potentially change the results generated by the code.  As some of these
changes may indeed resolve valid bugs or earlier problems with underlying
code, it will often be insufficient to demonstrate that results can be
reproduced when using the original versions, a problem sometimes known
as "code rot."  Researchers will want to know if the results are robust
to the changes. The case studies in [@Lapp_2014] provide examples of
these problems.

### 4. **Barriers to adoption and reuse in existing solutions**

Technological solutions such as workflow software, virtual machines,
continuous integration services, and best practices from software
development would address many of the issues frequently frustrating
reproducibility. However, researchers face significant barriers to entry
in learning these tools and approaches which are not part of their typical
curriculum, or lack incentives commensurate with the effort required
[@Joppa_2013; @FitzJohn_2014].

@@

<!-- A wide range of proposals, I focus on those at the operating system level -->
Though a wide variety of approaches exists to work around these challenges,
few operate on a low enough level to provide a general solution.  [@Clark_2014]
provide an excellent description of this situation:

> In scientific computing the environment was commonly managed via
Makefiles & Unix-y hacks, or alternatively with monolithic software
like Matlab. More recently, centralized package management has provided
curated tools that work well together. But as more and more essential
functionality is built out across a variety of systems and languages,
the value -- and also the difficulty -- of coordinating multiple tools
continues to increase. Whether we are producing research results or web
services, it is becoming increasingly essential to set up new languages,
libraries, databases, and more.

@@

<!-- FIXME Doesn't say much about the approaches. Not clear what a workflow is... -->

Two dominant paradigms have emerged to address these issues so far:
workflow software [e.g. @Altintas_2004; @Hull_2006]  and
cloud computing on virtual machines [e.g. @Dudley_2010; @Howe_2012]. 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from of
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they receive substantial adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall [@Gil_2007; @Dudley_2010].

[@Dudley_2010] gives several reasons that such comprehensive workflow
systems have not been more successful:

@@

In short, workflow software expects a new approach to computational research. 
In contrast, virtual machines (VMs) offer a more direct approach. Since
the computer Operating System (OS) already provides the software layer
responsible for coordinating all the different elements running on the computer,
the VM approach captures the OS and everything running on it whole-cloth.
To make this practical, [@Dudley_2010] and [@Howe_2012]
propose using virtual machine images that will run on the cloud, such as Amazon's
EC2 system, which is already based upon this kind of virtualization.

Critics of the use of VMs to support reproducibility highlight that the 
approach is too much of a black box and thus ill suited for reproducibility [@Watson_2012].
While the approach sidesteps the need to either install or even document the
dependencies, this also makes it more difficult for other researchers to
understand, evaluate, or alter those dependencies.  Moreover, other research
cannot easily build on the virtual machine in a consistent and scalable way. 
If each study provided it's own virtual machine, any pipeline combining the 
tools of multiple studies would quickly become impractical or impossible 
to implement.

@@

The problems highlighted here are not unique to _academic_ software,
but impact software development in general. While the academic research
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual
machines, the software development community has recently emphasized a
philosophy (rather than a particular tool), known as _Development_ and
Systems _Operation_, or more frequently just "DevOps." The approach is
characterized by scripting, rather than documenting, a description of the
necessary dependencies for software to run, usually from the Operating
System (OS) on up.  @Clark_2014 describes the DevOps approach along
with both its relevance to reproducible research and examples of its
use in the academic research context. They identify the difficulties
I have discussed so far in terms of effective documentation:

@@

The authors contrast this to the DevOps approach, where dependency 
documentation is _scripted_:

@@

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. @Clark_2014 are careful to note
that this is not so much a technological shift as a philosophical one:

@@

Nevertheless, a growing suite of tools designed explicitly for this
purpose have rapidly replaced the use of general purpose tools (such
as Makefiles, bash scripts) to become synonymous with the DevOps
philosophy.  @Clark_2014 reviews many of these DevOps tools, their
different roles, and their application in reproducible research.

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play
in reproducible research. Docker offers several promising features
for reproducibility that go beyond the tools highlighted in @Clark_2014.
Nevertheless, my goal in focusing on this technology is not to
promote a particular solution, but to anchor the discussion of technical
solutions to reproducibility challenges in concrete examples.

@@

Docker is an open source project that builds on many long-familiar
technologies from operating systems research: LXC containers,
virtualization of the OS, and a hash-based or git-like versioning
and differencing system, among others.

@@

A key difference between Docker images and other virtual machines is 
that the Docker images share the Linux kernel with the host machine.
For the end user the primary consequence of this is that any Docker
image must be based on a Linux system with Linux-compatible software, which
includes (R, Python, Matlab, and most other scientific programming needs). [^4]

Sharing the Linux kernel makes Docker much more light-weight and higher
performance than complete virtual machines -- a typical desktop computer
could run no more than a few virtual machines at once but 
would have no trouble running 100's of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

@@

Though Docker images can be created interactively, this leaves little
transparent record [^3] of what software has been installed and how.
Dockerfiles provide a simple script (similar to a Makefile) that defines
exactly how to build up the image, consistent with the DevOps approach
I mentioned previously.

With a syntax that is simpler than other provisioning tools (e.g. Chef,
Puppet, Ansible) or Continuous Integration (CI) platforms (e.g. Travis CI,
Shippable CI); users need little more than a basic familiarity with shell
scripts and a Linux distribution software environment (e.g. Debian-based
`apt-get`) to get started writing Dockerfiles.

@@

- the `Dockerfile` provides a human readable summary of the necessary
software dependencies, environmental variables and so forth needed to
execute the code. There is little possibility of the kind of holes or 
imprecision in such a script that so frequently cause difficulty in
manually implemented documentation of dependencies. This approach
also avoids the burden of having to tediously document dependencies 
at the end of a project, since they are instead documented as they 
are installed by writing the `Dockerfile`.

- Unlike a `Makefile` or other script, the `Dockerfile` includes all
software dependencies down to the level of the OS, and is built by the
Docker `build` tool, making it very unlikely that the resulting build
will differ when being built on different machines. This is not to say
that all builds of a Dockerfile are bitwise identical.  In particular,
builds executed later will install more recent versions of the same
software, if available, unless the package managers used are explicitly
configured otherwise. I address this issue in the next section.

- It is possible to add checks and tests following the commands for
installing the software environment, which will verify that the setup
has been successful.  This can be important in addressing the issue of
code-rot which I discuss next.

@@

[^3]: The situation is in fact slightly better than the virtual machine
approach because these changes are versioned. Docker provides tools
to inspect differences (diffs) between the images, and I can also roll
back changes to earlier versions.

@@

As I have discussed above, changes to the dependencies, whether they are
the result of security fixes, new features, or deprecation of old software,
can break otherwise functioning code. These challenges can be significantly
reduced because Docker defines the software environment to a particular operating system 
and suite of libraries, such as the Ubuntu or Debian distribution.  Such 
distributions use a staged release model with `stable`, `testing` and `unstable` 
phases subjected to extensive testing to catch such potential problems [@Ooms_2013],
while also providing regular security updates to software within each stage.
Nonetheless, this cannot completely avoid the challenge of code-rot,
particularly when it is necessary to install software that is not (yet)
available for a given distribution.

@@

<!-- FIXME move this later --> 
To simplify this process, Docker also supports *Automated Builds*
through the Docker Hub.  This acts as a kind of Continuous Integration
(CI) service that verifies the image builds correctly whenever the
Dockerfile is updated, particularly if the Dockerfile includes checks
for the environment. The Hub also provides a convenient distribution
service, freely storing the pre-built images, along with their metadata,
for download and reuse by others. The Docker Hub is a free service and
an open source software product so that users can run their own private
versions of the Hub on their own servers, for instance, if security of
the data or the longevity of the public platform is a concern.

@@

Though most of the concerns I have discussed so far can be addressed
through well-designed workflow software or the use of a DevOps approach
to provisioning virtual machines by scripts, neither approach has seen
widespread adoption by domain researchers, who work primarily in a
local rather than cloud-based environment using development tools native
to their personal operating system. To gain more widespread adoption,
reproducible research technologies must make it easier, not harder,
for a researcher to perform the tasks they are already doing (before 
considering any additional added benefits).

These issues are reflected both during the original research or development
phase and in any subsequent reuse. Another researcher may be less likely
to build on existing work if it can only be done by using a particular
workflow system or monolithic software platform with which they are
unfamiliar.  Likewise, a user is more likely to make their own computational
environment available for reuse if it does not involve a significant
added effort in packaging and documenting [@Stodden_2010].

Though Docker is not immune to these challenges, it offers a interesting 
example of a way forward in addressing these fundamental concerns.

@@

## Using Docker as a local development environment

Perhaps the most important feature of a reproducible research tool is
that it be easy to learn and fit relatively seamlessly into existing
workflow patterns of domain researchers.  This,
more than any other concern, can explain the relatively low uptake
of previously proposed solutions.  Being a new and unfamiliar tool to
most domain scientists, Docker is far from immune to the same critique.
Nevertheless, Docker takes us several key steps towards an approach that
can be easily adopted in a research context.

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [e.g. @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces are rapidly
filling this gap).

@@

For example, one launches an interactive R console in a container that
is linked to our current working directory like so:

```bash
docker run -v $(pwd):/ -it cboettig/rstudio /usr/bin/R
```

@@

That will take a while to download the image the first time you run
it. A boot2docker user will then need to determine the ip address assigned
to boot2docker, while a Linux user can just specify `http://localhost`:

@@

The key point here is that other researchers can easily build off this
image just created, extending our work directly, rather than having
to go back to the original image. Before others can re-use our image,
it must be made available. Docker provides a convenient way to do this
using the Docker Hub.

### Docker Hub
